Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-07-01 pharmaco.Rmd

_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the 4320 cases of severe COVID-19 in Scotland since the start of the epidemic.  Cases and controls were linked to records of hospital discharges in the last three years and dispensed prescriptions issued in primary care since June 2019.

@@

For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.

@@

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`. To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure. Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.  In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions. To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (subparagraph code 0407010).  The items grouped in this subparagraph include opiate-containing compound preparations.  `r paste.colpercent(with(subset(cc.severe, CASE==0), table(nonopioid.analgesic, opioid.analgesic)))[2, 2]` of those with a dispensed prescription of a drug in subparagraph 0407020 (opiate analgesics) also had a prescription in subparagraph 0207010.

To examine the associations with these drug classes in more depth, the total dispensed doses of proton pump inhibitors (as DDDs) of proton pump inhibitors and opiate (as MME, summed over subparagraphs 0407010 and 0407020) were calculated.  Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.  We used several approaches:

@@

## Associations with opiate use
Table \ref{tab:dosegr} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship, strongest in those aged less than 75 years.  With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with opiate dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations with proton pump inhibitor use
Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for COVID-19 increases with the dose of proton pump inhibitor.  The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile and any history of neoplasm.  This dose-response relationship was absent in those aged 75 years and over.  In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 2 DDDs/day MME /day were `r table.dosegr.protonpump.notlisted["2 or more", 4]` and `r table.dosegr.protonpump.notlisted["2 or more", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.

@@

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.

Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.  Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and ACE inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis. The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.

Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with drugs widely used in care homes: antipsychotics, analgesics and drugs for dementia.  Table \ref{tab:bnfchapter10} shows associations with drugs for the musculoskeletal system disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code. All antirheumatic drugs were associated with elevated rate ratios for COVID-19.  The univariate rate ratio associated with hydroxychloroquine sulfate was `r table.bnfchapter10[grep("Hydroxychloroquine", rownames(table.bnfchapter10)), 3]`; adjustment for ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.hydroxychlor.adjusted[1], coeff.hydroxychlor.adjusted[3])`.

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation. in individuals without conditions designated as conferring high risk.  The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017]. We attempted to investigate how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of specific drugs.  A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.  Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10, and that drug exposure is based on dispensed rather than issued prescriptions.

@@

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.  The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.  Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and more generally on the avoidance of unnecessary polypharmacy.

@@

## Conflicts of interest
HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly. HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.  HC and PM own stock in Roche and Bayer.

@@

knitr::kable(table.dosegr,
             escape=FALSE,
             label="dosegr", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average dose", colnames(table.dosegr.opiate)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of opiate (MME) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
	column_spec(column=1, width="3cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 16) %>%
    kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
    kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
    kableExtra::group_rows("Ages 0-74 years", 9, 13) %>%
    kableExtra::group_rows("Ages 75+ years", 14, 18) %>%
    landscape()

@@

colnames(table.timewindow.protonpump) <- colnames(table.timewindow.opiate)
table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opiate), 2),
                               table.timewindow)

knitr::kable(table.timewindow,
             escape=FALSE,
             label="timewindow", 
             booktabs=TRUE,
             row.names=FALSE, 
             col.names=c("Time window", clean.header(colnames(table.timewindow.opiate)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption=paste0("Associations of severe disease with dispensed prescriptions by ", TW, "-day time window")) %>%
	column_spec(column=1, width="3cm") %>%
	column_spec(column=2:3, width="2cm") %>%
	column_spec(column=4, width="2.5cm") %>%
	column_spec(column=5, width="1.5cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics", 1, 4) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
```

@@

knitr::kable(table.bnfchapter1,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter1",
             col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter2,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter2",
             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter4,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter4",
             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter10,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter10",
             col.names=c(clean.header(colnames(table.bnfchapter10)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
``````

>>>>

_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic.  Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care since June 2019.

@@

For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.

@@

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`. To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure. Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models with care home status and number of drug classes dispensed as covariates, in which each drug class was dropped from the total number of drug classes.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.  In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions. To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with two covariates -- number of cardiovascular drug classes and number of other drug classes -- the rate ratios per drug class for severe COVID-19 were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 50 cases and controls and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  A backward stepwise variable elimination procedure as described above was carried out to identify which of these `r length(subparacols.forstepwisedrop)` drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (BNF subparagraph code 0407010).

The items grouped under subparagraph 0407010 include opiate-containing compound preparations and also aspirin-containing preparations with which proton pump inhibitors are commonly co-prescribed. To examine the associations with these drug classes in more depth, the total dispensed daily dose of opiate as MME, summed over subparagraphs 0407010 and 0407020 (opiate analgesics) was calculated.  The total dispensed daily dose of proton pump inhibitors was calculated as DDDs.  Distinguishing between causality and confounding is challenging because opiate-containing drugs and proton pump inhibitores are usually co-prescribed with other drugs.  We used several approaches:

@@

## Associations with opiate use
Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship, strongest in those aged less than 75 years.  With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm. In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with opiate dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations with proton pump inhibitor use
Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for severe COVID-19 increases with the dose of proton pump inhibitor.  The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  This dose-response relationship was absent in those aged 75 years and over.  In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 2 DDDs/day MME /day were `r table.dosegr.protonpump.notlisted["2 or more", 4]` and `r table.dosegr.protonpump.notlisted["2 or more", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.

@@

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility and laxatives were strongly and independently associated with disease in a multivariate analysis.  Proton pump inhibitors were prescribed far more commonly than any other drug class in this chapter.

Table \ref{tab:bnfchapter2} shows associations with drugs for the cardiovascular system. Prescriptions of loop diuretics and anticoagulants were associated with elevated rate ratios for severe COVID-19 in univariate and multivariate analyses.  Over all age groups combined, the univariate rate ratio associated with oral anticoagulants was reduced from `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]` to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])` by adjustment for ischaemic heart disese, other heart disease and ever-use of a proton pump inhibitor.  Of drug classes commonly used to treat hypertension, ACE inhibitors and angiotensin II receptor antagonists were associated with reduced risk of COVID-19.

Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with antipsychotics, analgesics and drugs for dementia.  Table \ref{tab:bnfchapter10} shows associations with drugs for the musculoskeletal system disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code. All antirheumatic drugs were associated with elevated rate ratios for COVID-19.  The univariate rate ratio associated with hydroxychloroquine sulfate was `r table.bnfchapter10[grep("Hydroxychloroquine", rownames(table.bnfchapter10)), 3]`; adjustment for ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.hydroxychlor.adjusted[1], coeff.hydroxychlor.adjusted[3])`.

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation, in individuals without conditions designated as conferring high risk.  The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017]. We attempted to investigate how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of specific drugs.  A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.  Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10, and that drug exposure is based on dispensed rather than issued prescriptions.

@@

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing, althoug proton pump inhibitors and analgesics have the greatest explanatory power for the association of COVID-19 with polypharmacy.  Although this study did not identify unequivocal evidence of causality for opiates and proton pump inhibitors, effects of these drug classes on susceptibility to COVID-19 are biologically plausible.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of these drug classes, and more generally on the avoidance of unnecessary polypharmacy.

@@

## Conflicts of interest
HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly. HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.  HC and PM own stock in Roche and Bayer.  SH received honoraria from Gilead.

@@

knitr::kable(table.dosegr,
             escape=FALSE,
             label="dosegr", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average dose", colnames(table.dosegr.opiate)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of opiates and proton pump inhibitors, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
	column_spec(column=1, width="3cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 18) %>%
    kableExtra::group_rows("  Ages 0-74 years", 1, 4) %>%
    kableExtra::group_rows("  Ages 75+ years", 5, 8) %>%
    kableExtra::group_rows("  Ages 0-74 years", 9, 13) %>%
    kableExtra::group_rows("  Ages 75+ years", 14, 18) %>%
    landscape()

@@

colnames(table.timewindow.opiate) <- colnames(table.timewindow.protonpump)
table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opiate), 2),
                               table.timewindow)

knitr::kable(table.timewindow,
             escape=FALSE,
             label="timewindow", 
             booktabs=TRUE,
             row.names=FALSE, 
             col.names=c("Time window", clean.header(colnames(table.timewindow.protonpump)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption=paste0("Associations of severe disease with dispensed prescriptions by ", TW, "-day time window")) %>%
	column_spec(column=1, width="3cm") %>%
	column_spec(column=2:3, width="2cm") %>%
	column_spec(column=4, width="2.5cm") %>%
	column_spec(column=5, width="1.5cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics (restricted to those without history of neoplasm)", 1, 4) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
```

@@

knitr::kable(table.bnfchapter1,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter1",
             col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter2,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter2",
             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter4,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter4",
             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

@@

knitr::kable(table.bnfchapter10,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter10",
             col.names=c(clean.header(colnames(table.bnfchapter10)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
    column_spec(column=1, width="5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
``````
